---
title: "Reproducible Research: Peer Assessment 1"
author: "Eranda Ihalagedara"
date: "2023-08-12"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

```{r libraries, results=FALSE}
# Loading tidyverse and related libraries
library(tidyverse)
library(dplyr)
library(readr)
library(janitor)
library(ggplot2)

df = read_csv("activity.csv", na=c("NA"), col_names = TRUE)
```

After loading data, we can now examine data

```{r}
head(df)
```

```{r}
nrow(df)
```

```{r}
summary(df)
```

We can see that there are missing values in `steps` column

## What is mean total number of steps taken per day?

First we calculate total steps per each day

```{r}
df_per_day = df %>% 
  drop_na() %>% 
  group_by(date) %>% 
  summarise(total_steps = sum(steps))

head(df_per_day)
```

Histogram of the steps per day

```{r}
df_per_day %>% ggplot(aes(x=total_steps)) + 
  geom_histogram(binwidth = 1000, fill="aquamarine3") +
  labs(title="Steps per day", x="Steps", y="Count")
```

We can calculate the mean and the median

```{r}
df_per_day %>% 
  summarise(mean_steps_per_day = mean(total_steps), median_steps_per_day=median(total_steps))
```

## What is the average daily activity pattern?

Calculating the mean steps for each 5-minute interval across all days and plot

```{r}
df_per_interval = df %>% 
  drop_na() %>% 
  group_by(interval) %>% 
  summarise(average_steps =mean(steps))

df_per_interval %>% ggplot(aes(interval, average_steps)) +
  geom_line(color="blue3", size=.8) +
  labs(title="Average steps per interval", x="Interval", y="Average Steps")
```

The 5-minute interval with maximum average steps can be found as:

```{r}
df_per_interval %>% filter(average_steps==max(df_per_interval$average_steps))
```

## Imputing missing values

Total number of rows with missing values:

```{r}
df %>% is.na() %>% sum()
```

As we found out earlier in exploring data, all these missing values are from `steps`. An appropriate way of imputing missing steps would be using the medain of each 5 minute interval

```{r}
df_filled = df %>% 
  group_by(interval) %>% 
  mutate(steps = if_else(is.na(steps), 
                            median(steps, na.rm = TRUE), 
                            steps))

head(df_filled)
```

Histogram of steps per day with missing data filled dataset

```{r}
df_filled_day = df_filled %>% 
  group_by(date) %>% 
  summarise(total_steps = sum(steps))

df_filled_day %>% ggplot(aes(total_steps)) +
  geom_histogram(binwidth = 1000, fill="aquamarine3")

```

The mean and the median

```{r}
df_filled_day %>% 
  summarise(mean_steps_per_day = mean(total_steps), median_steps_per_day=median(total_steps))
```

**From the histogram and the mean/median statistics we can see the dataset with imputed missing values differ from the original dataset. There is a increase in the count of the histogram in the lower end. Further, mean and the median has decreased that the original dataset, indicating the imputed values are low value and in greater count.**

## Are there differences in activity patterns between weekdays and weekends?

Add a new column to indicate weekdays and weekends

```{r}
df_filled_day = df_filled %>% 
  mutate(day=if_else(weekdays(date)!="Saturday" & (weekdays(date)!="Sunday"),"weekday","weekend")) %>% 
  mutate(day=as.factor(day))

head(df_filled_day)
```

```{r}
df_filled_day %>% 
  group_by(interval, day) %>% 
  summarize(avg_steps = mean(steps)) %>% 
  ggplot(aes(interval, avg_steps)) +
  geom_line(color="blue3", size=.8) +
  facet_wrap(~day) +
  labs(title="Average steps per interval by day", x="Interval", y="Average Steps")
```

**According to the plots, during weekdays average steps in the mid-day hours are lower than that of weekend. Also there is higher peak in the early hours during weekdays**
